### 2011-05-09 draft2.tex
Parser for a toy haskell language and updates about the program translation.

\thispagestyle{empty}
\section{Grammars}

@@

\subsection{$\lambda$-lifted haskell subset}
\begin{center}
\begin{array}{rclr}
  u,e &:=& x \mid f \mid e~e \mid D(e,\dots,e) \mid \bad \mid n & (Expression)\\
  p &:=& \Delta,p \mid T,p \mid \epsilon & (Program)\\
  %\Delta &:=& d \mid transp (d,c) \mid opaque (d,c) & (Defintion~and~contract)\\
  d &:=& f~x_1 \dots x_n = e \mid f~x_1 \dots x_n = \mbox{ case } e \mbox{ of } [(pat_i,e_i)] & (Definition)\\
  T &:=& \mbox{data } x~\tau_1 \dots \tau_n = D_1 ; D_2 ; \dots ; D_n & (Data~type~definition)\\ %TODO make type clearer
  pat &:=& D(x_1,\dots,x_n) & (Pattern)\\
\end{array}
\end{center}

@@

\section{Translation}
We define several translations: $\etrans{},\dtrans{},\ttrans{},\strans{}$.
\begin{center}
\begin{array}{rclr}
  \etrans{} &::& Expression \to Term\\
  \dtrans{} &::& Definition \to FOF\\
  \ttrans{} &::& Data~type  \to FOF\\
  \strans{} &::& Expression \to Contract \to FOF\\
%  \trans {} &::& Definition \to Contract \to FOF\\
\end{array}
\end{center}

\subsection{$\etrans{}$}
$\etrans{e}$ is a term.

\begin{eqnarray}
\etrans{x} &=& x\\
\etrans{f} &=& f\\
\etrans{e_1~e_2} &=& app(e_1,e_2)\\
\etrans{D(e_1,\dots,e_n)} &=& D(\etrans{e_1},\dots,\etrans{e_n}\\
\etrans{\bad} &=& \bad\\
\etrans{n} &=& n\\

@@

\begin{eqnarray}
\dtrans{f~x_1 \dots x_n = e} &=& \forall x_1 \dots x_n.\etrans{f~x_1 \dots x_n} = \etrans{e}\\
\dtrans{f~x_1 \dots x_n = \mbox{case } e \mbox{ of } [D_i~\overline{z} \mapsto e_i]} &=& \forall x_1 \dots x_n~(\bigwedge_i (\forall \overline{z}~\etrans{e} = \etrans{D_i~\overline{z}} \to \etrans{f~x_1 \dots x_n} = \etrans{e_i})\\
&& \land \etrans{e} = \bad \to \etrans{f~x_1 \dots x_n} = \bad)\\ 
&& \land (\etrans{f~x_1 \dots x_n} = \unr \bigvee_i (\texttt{HD}(e) = D_i) \lor e = \bad)
\end{eqnarray}

Can't we just put a XOR instead of OR in the last line?

@@

\thispagestyle{empty}

\section{Datatypes}
$\ttrans{T}$ is a set of first-order formulae.
$\ttrans{\mbox{data } T = D_1 ; \dots ; D_n} = S_1 \cup S_2 \cup S_3$.

First, for each $D_i$ we introduce selectors $sel_{k,D_i}$, which are basically projections of $D_i(x_1,\dots,x_n)$ on its $k$-th composant, so that we can express that constructors are injective :
$$S_1 := \{ \forall x_1,\dots,x_n . \bigwedge_{1 \leq i \leq n} sel_{i,D_i}(D_i(x_1,\dots,x_n)) = x_i \mid 1 \leq i \ n \}$$

For each pair of different constructors $D_i,D_j$, we state that they can never map to the same value: 
$$S_2 := \{ \forall x_1,\dots,x_n~\forall y_1,\dots,y_k . D_i(x_1,\dots,x_n) \neq D_j(y_1,\dots,y_k) \mid 1 \leq i < j \leq n \}$$

Finally, we have to give crash-freeness conditions for each $D_i$:
$$S_3 := \{ \forall x_1,\dots,x_n . (\cf{x_1} \land \dots \land \cf{x_n} \leftrightarrow \cf{D_i(x_1,\dots,x_n)}) \mid 1 \leq i \leq n \}$$

\thispagestyle{empty}
\end{document}

>>>>

\section{Grammars}

@@

\subsection{$\lambda$-lifted haskell subset}
\begin{center}
\begin{array}{rclr}
  u,e &:=& x \mid f \mid e~e \mid D(e,\dots,e) \mid \bad \mid n & (Expression)\\
  p &:=& \Delta,p \mid T,p \mid f \in c,p \mid \epsilon & (Program)\\
  \Delta &:=& d \mid opaque(d) & (Defintion~and~scope)\\
  d &:=& f~x_1 \dots x_n = e \mid f~x_1 \dots x_n = \mbox{ case } e \mbox{ of } [(pat_i,e_i)] & (Definition)\\
  T &:=& \mbox{data } x = D_1 ; D_2 ; \dots ; D_n & (Data~type~definition)\\ %TODO make type clearer
  pat &:=& D(x_1,\dots,x_n) & (Pattern)\\
\end{array}
\end{center}

For the moment we consider data constructors to be saturated, ie fully applied (hence the special application syntax).

@@

\section{Translation}
We define several translations: $\etrans{},\dtrans{},\ttrans{},\strans{},\trans{}$.
\begin{center}
\begin{array}{rclr}
  \etrans{} &::& Expression \to Term\\
  \dtrans{} &::& Definition \to FOF\\
  \ttrans{} &::& Data~type  \to FOF\\
  \strans{} &::& Expression \to Contract \to FOF\\
  \trans {} &::& Program    \to \{ FOF \}\\
\end{array}
\end{center}

\subsection{$\etrans{}$}
$\etrans{e}$ is a term, the translation is really straightforward.

\begin{eqnarray}
\etrans{x} &=& x\\
\etrans{f} &=& f\\
\etrans{e_1~e_2} &=& app(e_1,e_2)\\
\etrans{D(e_1,\dots,e_n)} &=& D(\etrans{e_1},\dots,\etrans{e_n})\\
\etrans{\bad} &=& \bad\\
\etrans{n} &=& n\\

\subsection{$\ttrans{}$}
$\ttrans{T}$ is a set of first-order formulae which we break down in three parts:
$\ttrans{\mbox{data } T = D_1 ; \dots ; D_n} = S_1 \cup S_2 \cup S_3$.

First, for each $D_i$ of arity $n_i$ we introduce selectors $sel_{k,D_i}$, which are projections of $D_i(x_1,\dots,x_n_i)$ on its $k$-th composant, so that we can express that constructors are injective :
$$S_1 := \{ \forall x_1,\dots,x_{n_i} . \bigwedge_{1 \leq j \leq n} sel_{j,D_i}(D_i(x_1,\dots,x_{n_i})) = x_j \mid 1 \leq i \leq n \}$$

For each pair of different constructors $D_i,D_j$, we state that they can never map to the same value: 
$$S_2 := \{ \forall x_1,\dots,x_{n_i}~\forall y_1,\dots,y_{n_j} . D_i(x_1,\dots,x_{n_i}) \neq D_j(y_1,\dots,y_{n_j}) \mid 1 \leq i < j \leq n \}$$

Finally, we have to give crash-freeness conditions for each $D_i$:
$$S_3 := \{ \forall x_1,\dots,x_{n_i} . (\cf{x_1} \land \dots \land \cf{x_{n_i}} \leftrightarrow \cf{D_i(x_1,\dots,x_{n_i})}) \mid 1 \leq i \leq n \}$$

@@

\begin{eqnarray}
\dtrans{f~\overline{x} = e} &=& \forall x_1 \dots x_n.\etrans{f~x_1 \dots x_n} = \etrans{e}\\
\dtrans{f~\overline{x} = \mbox{case } e \mbox{ of } [D_i(\overline{z}) \mapsto e_i]} &=& \forall x_1 \dots x_n.(\bigwedge_i (\forall \overline{z}~\etrans{e} = \etrans{D_i(\overline{z})} \to \etrans{f~\overline{x}} = \etrans{e_i})\\
&& \land \etrans{e} = \bad \to \etrans{f~x_1 \dots x_n} = \bad)\\ 
&& \land ((\bigwedge_i e \neq D_i(sel_{1,D_i}(e),\dots,sel_{n_i,D_i}(e)) \land \etrans{e} \neq \bad)\\
&& ~~\to \etrans{f~x_1 \dots x_n} = \unr)
\end{eqnarray}

An alternative to using selectors would be to write something along the line of $\exists y_1,\dots,y_{D_i}. e \neq D_i(y_1,\dots,y_{D_i})$. It is equivalent but the skolemisation process of equinox will anyway turn those existentials in selectors functions. But if we pattern match $D_i$ in two different functions, we would get two different selectors (the same selector with two different names) so it's better (for the efficiency of the proof) to define selectors for each $D_i$ once in for all and use it as much as possible afterwards.

@@

\subsection{$\trans{}$}
$\trans{p}$ defines the translation of a program to a theory (a set of FO formulae)

\begin{eqnarray}
\trans{\epsilon} &=& \emptyset\\
\trans{d,p} &=& \dtrans{d} \cup \trans{p}\\
\trans{opaque(d),p} &=& \dtrans{d} \cup \trans{p}\\
\trans{T,p} &=& \ttrans{T} \cup \trans{p}\\
\trans{f \in c,p} &=& \strans{f \in c} \cup \trans{p}\\
\end{eqnarray}

\section{User interaction}

The user provides a program $p$ which consists of functions definition (either opaque or transparent), data types definition and claims that functions satisfies contracts.

Then, for each function definition $d_f := f~x_1 \dots x_n = e$ of a function $f$ that has to satisfy the set of contracts $C_f$, we construct the context $C = p \setminus ( d_f \cup C_f )$, which is basically the program $p$ without the definition of $f$ and the contracts it has to satisfy.

We then want to check (with equinox) that:
 $$\trans{C} \cup \{ \dtrans{f~x_1 \dots \x_n = e[f^\star/f]}, \bigwedge_{c \in C_f} \strans{f^\star \in c} \} \models \bigwedge_{c \in C_f} \strans{f \in c}$$

Which we rewrite as:
 $$\trans{C} \cup \{ \dtrans{f~x_1 \dots \x_n = e[f^\star/f]}, \bigwedge_{c \in C_f} \strans{f^\star \in c},  \bigvee_{c \in C_f} \lnot \strans{f \in c} \} \models \bot$$

\section{Remarks}
\begin{itemize}
\item What to do with mutually recursive functions?
\item Explore the variants translations (cf variants of the popl paper)
\end{itemize}

\end{document}
